---
title: "STAT 639V - Assignment 1"
author: "Carson Stacy"
date: "9/1/2021"
output:
  pdf_document
---


**1. Let Y be a discrete random variable taking values from {1, 2, 3} with probabilities {0.2, 0.4, 0.4}. A sample of Y is {3, 2, 3, 1, 2, 2, 2, 3, 1}. Calculate the sample mean, sample variance, theoretical mean and theoretical variance for Y. (10 points)**

given: $n = 9 , y = \{3, 2, 3, 1, 2, 2, 2, 3, 1\} ,  P(1) = 0.2 , P(2) = 0.4 , P(3) = 0.4$

$\underline{Sample\space mean:}$ $$\bar{y} = \frac{1}{n} * \sum_{i = 1}^{n} y_{i} = \frac{3 + 2 + 3 + 1 + 2 + 2 + 2 + 3 + 1}{9} = \frac{19}{9} = \boxed{2.\bar{1}}$$

$\underline{Sample\space variance:}$ $$S^2 = \frac{1}{n-1} * \sum_{i = 1}^{n} (y_{i} - \bar{y})^2 = \frac{3(3 - 2.\bar{1})^2 + 4(2 - 2.\bar{1})^2 + 2(1 - 2.\bar{1})^2}{9-1} \approx \frac{2.37 + 0.05 + 2.47}{8} \approx \boxed{0.611}$$


$\underline{Theoretical\space mean:}$ $$\mu = E(Y) = \sum_{j = 1}^{k} y_{j}p_{j} = (1 * 0.2) + (2 * 0.4) + (3 * 0.4) = \boxed{2.2}$$


$\underline{Theoretical\space variance:}$ $$\sigma^2 = E(Y - \mu)^2 = \sum_{j = 1}^{k} (y_{j} - \mu)^2p_{j} = (1 - 2.2)^2 * 0.2 +  (2 - 2.2)^2 * 0.4 + (3 - 2.2)^2 * 0.4 = \boxed{0.56}$$

\pagebreak


**2. The 2012 Gallup survey suggests that 26.2% of Americans are obese. Among a random sample of 10 Americans, what is the probability that (a) at least 2 are obese? (b) no more than 7 are obese? (10 points)**

given: $$n = 10,\space p=0.262,\space P(k\space success\space in\space n\space trials) = \binom{n}{k}p^k(1-p)^{(n-k)}$$

(a) The probability that at least 2 are obese $P(k \ge 2)$ is equivalent to $1 - P(k < 2)$ which is equivalent to $P(k \ge 2) = 1 - [P(k=0) + P(k=1)]$

We can find that

$$P(k=0) = \binom{n}{k}p^k(1-p)^{(n-k)} = \binom{10}{0}0.262^0(1-0.262)^{(10-0)} \approx 0.0479$$
and 
$$P(k=1) = \binom{n}{k}p^k(1-p)^{(n-k)} = \binom{10}{1}0.262^1(1-0.262)^{(10-1)} \approx 0.1701$$
which can then be used to calculate $P(k \ge 2)$:
$$P(k \ge 2) = 1 - [P(k=0) + P(k=1)] \approx 1 - (0.0479 + 0.1701) \approx \boxed{0.7819}$$
which can be confirmed in R code:
```{r}
pbinom(1,10,0.262, lower.tail = FALSE)
```


(b) The probability that no more than 7 are obese $P(k \le 7)$ is equivalent to $1 - P(k > 7)$ which is equivalent to $P(k \le 7) = 1 - [P(k=8) + P(k=9)+ P(k=10)]$


We can find that

$$P(k=8) = \binom{n}{k}p^k(1-p)^{(n-k)} = \binom{10}{8}0.262^8(1-0.262)^{(10-8)} \approx 0.0005441712$$
and 
$$P(k=9) = \binom{n}{k}p^k(1-p)^{(n-k)} = \binom{10}{9}0.262^9(1-0.262)^{(10-9)} \approx 0.0000429307$$
and 
$$P(k=10) = \binom{n}{k}p^k(1-p)^{(n-k)} = \binom{10}{10}0.262^{10}(1-0.262)^{(10-10)} \approx 0.000001524$$

which can then be used to calculate $P(k \le 7)$:
$$P(k \le 7) = 1 - [P(k=8) + P(k=9)+ P(k=10)] \approx 1 - (0.0005441712 + 0.0000429307 + 0.000001524) \approx \boxed{0.9994114}$$

which can be confirmed in R code:
```{r}
pbinom(7,10,0.262, lower.tail = TRUE)
```



**3. A bowl contains 10 marbles including 2 red marbles, 3 green marbles, and 5 blue marbles. If we randomly select 4 marbles from the bowl, WITH replacement, what is the probability of selecting 2 green marbles and 2 blue marbles? (10 points)**

given: $n = 4,\space n_{red}=2,\space n_{green} = 3,\space n_{blue}=5,\space P(green) = 0.3,\space P(red)=0.2,\space P(blue)=0.5$

According to the multinomial formula,
$$P(n_{green}=2, n_{blue}=2, n_{red}=0) = \frac{n!}{n_{green}!n_{blue}!n_{red}!} * P(green)^{n_{green}} * P(blue)^{n_{blue}} * P(red)^{n_{red}} $$
$$=  \frac{4!}{2!2!0!} * 0.3^{2} * 0.5^{2} * 0.2^{0} = 6 * 0.09 * 0.25 * 1 = \boxed{0.135}$$
\
\
\

**4. Suppose X $\sim$ Pois($\lambda 1$) and Y $\sim$ Pois($\lambda 2$) are independent Poisson random variables. If $\lambda 1$ = 2 and $\lambda 2$ = 3, calculate E(X - Y) and Pr(X + Y = 1). (10 points)**

Given $X \sim {\sf Pois}(\lambda_1)$ and $Y \sim {\sf Pois}(\lambda_2)$ where $\lambda_1 = 2$ and $\lambda_2 = 3$, the expectation of the difference of X and Y, $E(X - Y)$ is equal to $E(X) - E(Y)$ because the random variables are independent. For random variables following a poisson distribution, the $E(A) = \lambda_A$. In this case, $$E(X - Y) = E(X) - E(Y) = \lambda_1 - \lambda_2 = 2 - 3 = \boxed{-1}$$ While $\lambda$ of a poisson process is greater than zero $\lambda \in (0,\infty)$, the difference between two poisson processes, termed a Skellam Distribution, has a domain that includes negative integers, therefore I believe this answer is correct.

The $Pr(X+Y = 1)$ can be found according to the Probability Mass Function (PMF) for poisson processes below, where $j=1$ and lambda of the linear combination of independent random variables X and Y of X + Y results in a $\lambda = 5$ because $E(X + Y) = E(X) + E(Y) = \lambda_1 + \lambda_2 = 2 + 3 = 5$

Poisson PMF
$$P(X + Y = j) = \frac{\lambda^j}{j!}* e^{-\lambda}$$

Given that $j=1$ and $\lambda = 5$ for the linear combination X + Y, then
$$P(X + Y = 1) = \frac{5^1}{1!}* e^{-5} = 5e^{-5} \approx \boxed{0.03368}$$
\pagebreak

**5. Suppose random variable Y follows a normal distribution N($\mu$ = 0, $\sigma ^ 2$ = 4). Calculate the density at Y=2, i.e., f(2). Use R or table to calculate the inter-quartile range, Q.75 - Q.25.(10 points)**

(a) Given $X \sim N(\mu = 0, \sigma^2 = 4)$, we know that $\sigma = \sqrt{4} = 2$.

From the PMF for the normal distribution
$$P(Y=y) = f(y) = \frac{1}{\sigma\sqrt{2\pi}}e^{-\frac{(y-\mu)^2}{2\sigma^2}}$$
we can find $P(y = 2)$:
$$P(Y=2) = f(2) = \frac{1}{2\sqrt{2\pi}}e^{-\frac{(2-0)^2}{2*4}} \approx \boxed{0.1209854}$$
which can be confirmed in R:
```{r}
dnorm(2, mean = 0, sd = 2)
```

(b) The IQR of $X \sim N(\mu = 0, \sigma^2 = 4)$ can be determined by subtracting $Q_{0.75} - Q_{0.25}$, done here with using the qnorm function in R:
```{r}
qnorm(p = 0.75, mean = 0, sd = 2) - qnorm(p = 0.25, mean = 0, sd = 2)
```
We can also see this via simulation
```{r}
# set seed for reproducability
set.seed(1)
# create simulated dataset with mu=0 and sd=2, save as y
y <- rnorm(1000000, mean = 0, sd = 2)
#find the iqr for y 
IQR(y)

# should be close to the theoretical IQR determined above..
```

$$ IQR \approx \boxed{2.697959} $$
\pagebreak

**6. A coin was flipped 10 times independently, and 3 heads were observed. Let p be the probability of getting a head, compute the maximum likelihood estimate of p. (10 points)**

given: $n=10,\space y = 3,\space p=P(H)$

$$Y \sim {\sf Bin}(n, p)$$
The maximum likelihood estimate for p ($\hat{p}_{MLE}$) can be calculated according to $\hat{p}_{MLE} = \frac{y}{n}$ where n is the number of trials and y is the number of successes.

By the binomial proportion equation,
$$\hat{p}_{MLE} = \boxed{\frac{3}{10}}$$

