---
title: "STAT 639V - Assignment 2"
author: "Carson Stacy"
date: "9/30/2021"
output:
  pdf_document
---

```{r include=FALSE}
require(tidyverse)
require(ggpubr)
```

**1. Let $X$ have a uniform distribution on 0 to 100 days with probability density function $f(x) = 1/100$ for $0 < x < 100$. (10 points)**


```{r echo=FALSE, fig.height=2, fig.width=2}
x <- c(1:100)
y <- dunif(x, 0, 100) #rep(0.01,101)
z <- punif(x, 0, 100)
S <- 1-z
df <- data.frame(x,y,z,S)



p1 <- ggplot(data = df, aes(x=x, y=y)) +
  geom_line() +
  ylim(0, 0.1) +
  labs(x = "x", y = "f(x)") +
  ggtitle("pdf for X") +
  theme_bw()
p2 <- ggplot(data = df, aes(x=x, y=z)) +
  geom_line() +
  ylim(0, 1) +
  labs(x = "x", y = "F(x)") +
  ggtitle("cdf for X") +
  theme_bw()
p3 <- ggplot(data = df, aes(x=x, y=S)) +
  geom_line() +
  ylim(0, 1) +
  labs(x = "x", y = "S(x)") +
  ggtitle("survival for X") +
  theme_bw()

p1
p2
p3

```

**(a) Find the survival function at 25, 50 and 75 days.**
$$S(x) = 1 - F(x)$$

The survival function at 25 days:
$$S(X=25) = 1 - F(X=25) = 1 - P(X \leq 25) = 1 - \frac{25 - 1+ 1}{100} = 1- \frac{25}{100} = \boxed{0.75} $$
The survival function at 50 days:
$$S(X=50) = 1 - F(X=50) = 1 - P(X \leq 50) = 1 - \frac{50 - 1+ 1}{100} = 1- \frac{50}{100} = \boxed{0.50} $$

The survival function at 75 days:
$$S(X=75) = 1 - F(X=75) = 1 - P(X \leq 75) = 1 - \frac{75 - 1+ 1}{100} = 1- \frac{75}{100} = \boxed{0.25} $$

**(b) Find the mean residual lifetime at 25,
50 and 75 days.**

We know that $$mrl(X=k) = \frac{\int_{k}^{\infty}(t-k)\times f(t)dt}{S(X=k)}$$

The mean residual lifetime at 25 days:
$$mrl(X=25) = \frac{\int_{25}^{100}(t-25)\times\frac{1}{100}dt}{\frac{3}{4}} = 
\frac{1}{75}\int_{25}^{100}(t-25)dt = 
\frac{1}{75}(\frac{1}{2}t^2\mid_{25}^{100} - 25t\vert_{25}^{100}) $$
$$=
\frac{1}{75}(\frac{1}{2}\times100^2 - \frac{1}{2}\times25^2 - 25\times100 + 25\times25) = 37\frac{1}{2} = \boxed{37.5} days
$$


The mean residual lifetime at 50 days:

$$mrl(X=50) = \frac{\int_{50}^{100}(t-50)\times\frac{1}{100}dt}{\frac{1}{2}} = 
\frac{1}{50}\int_{50}^{100}(t-50)dt = 
\frac{1}{50}(\frac{1}{2}t^2\mid_{50}^{100} - 50t\vert_{50}^{100}) $$
$$=
\frac{1}{50}(\frac{1}{2}\times100^2 - \frac{1}{2}\times50^2 - 50\times100 + 50\times50)= \boxed{25} days$$

The mean residual lifetime at 75 days:

$$mrl(X=75) = \frac{\int_{75}^{100}(t-75)\times\frac{1}{100}dt}{\frac{1}{4}} = 
\frac{1}{25}\int_{75}^{100}(t-25)dt = 
\frac{1}{25}(\frac{1}{2}t^2\mid_{75}^{100} - 75t\vert_{75}^{100}) $$
$$=
\frac{1}{25}(\frac{1}{2}\times100^2 - \frac{1}{2}\times75^2 - 75\times100 + 75\times75) =  \boxed{12.5} days$$




<!-- \pagebreak -->


**2. Let $X$ have a uniform distribution on [0,$\theta$], find the hazard function and mean residual life function of $X$.
(10 points)**

given: $h(x) = \frac{f(x)}{S(x)}$. Assuming that $X$ is continuous, we know that $f(x) = \frac{1}{\theta - 0} = \frac{1}{\theta}$ and $S(x) = 1 - \frac{x}{\theta} = \frac{\theta - x}{\theta}$ for $x \in [0,\theta]$. 

Therefore,
$h(x) = \frac{\frac{1}{\theta}}{\frac{\theta - x}{\theta}} = \frac{1}{\theta} \times \frac{\theta}{\theta-x} = \boxed{\frac{1}{\theta-x}}$


To calculate the mean residulal life for a given $x$, we have:

$$mrl(x) = \frac{\int_x^\infty(t-x)f(t)dt}{S(x)} = \frac{\int_x^\theta(t-x)\frac{1}{\theta}dt}{\frac{\theta - x}{\theta}} = 
\frac{\frac{1}{\theta}\int_x^\theta(t-x)dt}{\frac{\theta - x}{\theta}}
$$

$$= \frac{\frac{1}{\theta}(\frac{1}{2}t^2\mid_{t}^{\theta} - xt\vert_{t}^{\theta})}{\frac{\theta - x}{\theta}}
= \frac{1}{\theta-x}(\frac{1}{2}t^2\mid_{t}^{\theta} - xt\vert_{t}^{\theta})$$
$$= \frac{1}{\theta-x}(\frac{1}{2} \times \theta^2 - \frac{1}{2} \times x^2 - \theta x + x^2)
= \frac{1}{\theta-x}(\frac{\theta^2}{2} + \frac{x^2}{2} - \theta x)
= \frac{1}{2}\frac{1}{\theta-x}(\theta - x)^2 = \boxed{\frac{\theta - x}{2}}$$


**3. The lifetime of light bulbs follows an exponential distribution with a hazard rate of 0.001 failures per hour of use ($\lambda$ = 0.001). (10 points)
(a) Find the mean lifetime of a randomly selected light bulb.**


The mean lifetime of a randomly selected lightbulb whose lifetime follows an exponential distribution with $h(x) = \lambda= 0.001$ is the expectation of the r.v. X, the lifetime of lightbulbs. 


Because $\lambda = \frac{1}{1000}$, the $E(X) = \frac{1}{\lambda} = \boxed{1000}$ This is a characteristic of the expectation of an exponential distributuion.



**(b) What is the probability a light bulb will still function after 2,000 hours of use? **

From the hazard function, we know that $$S(x) = e^{-\int_0^x \frac{1}{1000}du} = e^{\frac{-x}{1000}}$$

Solving the survival function at X = 2000, we have:

$$S(X = 2000) = exp(\frac{-2000}{1000} = e^{-2} \approx \boxed{0.135} )$$

which means the probability a light bulb will survive over 2,000 hours of use is about 0.135.



<!-- Because $f(x) = - (S'(x))$, we find the p.d.f. of X to be $$\frac{1}{1000} \times exp(\frac{-x}{1000})$$ where $x > 0$. -->

<!-- The p -->

\pagebreak

**4. The time in days to development of a tumor for rats exposed to a carcinogen follows a Weibull distribution with $\alpha$ = 2 and $\lambda$ = 0.001. (10 points)**

**(a) What is the probability a rat will be tumor free at 30 days?** 

Let X be the time to tumor development in days where $X$ ~ $Weibull(\alpha = 2, \lambda = 0.001)$

The p.d.f. of the weibull distribution is $$f(x) = \alpha\lambda e^{-\lambda x^\alpha} \times x^{\alpha-1} = \frac{2}{1000}e^{\frac{-x^2}{1000}} \times x^{2-1} = \frac{x}{500} e^{\frac{-x^2}{1000}}$$

From this p.d.f., the $S(t)$ or $P(X>t)$ can be found by integration from $t$ to $\infty$:

$$S(t) = P(X>t) = \int_t^{\infty} \frac{x}{500} e^{\frac{-x^2}{1000}}dx = -e^{\frac{-x^2}{1000}}\vert_{t}^{\infty} = e^{\frac{-t^2}{1000}}$$

Setting $ t = 30$, we find:
$$P(X>30) = e^{\frac{-30^2}{1000}} = e^{-0.9} \approx \boxed{0.407}$$

**(b) What is the mean time to tumor?** ($\Gamma(0.5) = \sqrt{\pi}$).

We know that for the Weibull distribution, the $$E(X) = \Gamma(1 + \frac{1}{\alpha}) \times \lambda^{\frac{-1}{\alpha}}$$

so $$E(X) = \Gamma(1 + \frac{1}{2}) \times (\frac{1}{1000})^{-\frac{1}{2}} = \Gamma(\frac{3}{2}) \times \sqrt{1000} = \frac{1}{2}\sqrt{\pi} \times \sqrt{1000} \approx \boxed{28.02} days$$


**(c) Find the median time to tumor.**

Let $S(x) = 0.5$ and solve for x to find the median time to tumor: $$S(t) = 0.5 = e^\frac{-t^2}{1000}$$

$$log(0.5) = \frac{-t^2}{1000}$$
$$\sqrt{-1000 \times log(0.5) } = t \approx \boxed{26.33}$$
